---
title: "Introduction to R: practical 1.1"
author: "PBRC: DA Hughes"
date: "Oct 9th, 2024"
output: html_document
---

```{r render, eval=F, echo=F}
## install.packages("rmarkdown")
library(rmarkdown)

show.solutions <- F
render("practical.rmd",output_file="practical.html")

show.solutions <- T
render("practical.rmd",output_file="solutions.html")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

if (!exists("show.solutions"))
    show.solutions <- F
## show output for solutions copy only, use as results=hide.results()
hide.results <- function() ifelse(show.solutions, "markup", "hide")
## show code and output for solutions copy only, use as include=hide.all()
hide.all <- function() show.solutions
## show figure for solutions copy only, use as show.fig=hide.figure()
hide.figure <- function() ifelse(show.solutions, "asis", "hide")
```

## Objectives

By the end of this practical it is intended that you will be able to:

* Define variables
  * vectors
  * matrixes
  * data frames
  * lists
* Perform math with variables 
* Subset variables

Begin by opening R Console or a Terminal and starting an R session. Work directly in the interactive R console. 


## Defining variables and simple addition 

### Vectors

1. Define the variable '**x**' as a vector with the values 1 to 10.
```{r}
x = 1:10
```

2. Define a second variable '**y**' as a vector with the value 10. 
```{r}
y = 10
```

3. Add '**x**' to '**y**'.
```{r, results=hide.results()}
x + y
```

4. What is happening with the value assigned to the variable '**y**'?

--

5. Redefine '**y**' the values 10 and 20. 
```{r}
y = c(10,20)
```

6. Now add '**x**' to '**y**'.
```{r, results=hide.results()}
x + y
```

7. What is happening with the values assigned to the variable '**y**'?

--

8. Redefine '**y**' the values 10, 20, and 30. 
```{r}
y = c(10, 20, 30)
```

9. Now add '**x**' to '**y**'.
```{r, results=hide.results(), warning=FALSE}
x + y
```

10. What is happening with the values assigned to the variable '**y**'?

--

11. What were to happen if the variable '**x**' had the values 1, 2, and 3?
```{r, results=hide.results()}
x = 1:3
x + y
```

11. What were to happen if the variable '**x**' had the values 1, 2, and "PBRC"?
```{r, results=hide.results(), error=TRUE, warning=FALSE, message=FALSE}
x = c(1:2, "PBRC")
# x + y
```

12. Could you force the addition of the first and second elements of x to the value 10?

```{r, results=hide.results()}
as.numeric(x[1:2]) + 10
```

### Matrixes

13. Define a matrix '**mat**' with 3 rows and 3 columns with the values 1 to 9

```{r}
mat = matrix(1:9, nrow = 3, ncol = 3)
```

14. Add the value 100 to all elements of the matrix

```{r, results=hide.results()}
mat + 100
```
15. Add the value 100 to only the 2nd column of the matrix '**mat**'

```{r, results=hide.results()}
mat[,2] + 100
```

15. Add the value 100 to only the 3rd row of the matrix '**mat**'

```{r, results=hide.results()}
mat[3,] + 100
```

16. Give the matrix '**mat**' column names "cat", "dog", and "donkey".

```{r, results=hide.results()}
colnames(mat) = c("cat", "dog", "donkey")
mat
```
17. Multiply the column donkey to the column dog

```{r, results=hide.results()}
mat[, "dog"] + mat[, "donkey"]
```

### Data Frames

18. Define a data frame '**df**' with three columns called "ID", "height" (m), and "weight" (kg) with 3 observations each

```{r}
df = data.frame(ID = paste0(rep("ID_", 3) , 1:3), height = c(1.82, 1.73, 1.67), weight = c(90, 110, 85))
```

19. Calculate BMI (kg/m2) from the data frame you just made

```{r, results=hide.results()}
(BMI = df$weight / df$height^2)
# BMI
```

20. Add BMI to the data frame

```{r, results=hide.results()}
df$BMI = BMI
df
```

### Lists

21. Place the variables '**y**', '**mat**', and '**df**' into a list element called '**mylist**'.

```{r}
mylist = list(y = y, mat = mat, df = df)
```

22. Using the list add the '**mat**' column *donkey* to the '**df**' column *BMI*. 

```{r, results=hide.results()}
mylist$mat[, "donkey"] + mylist$df$BMI
```

23. Is there a second way to do the same thing? Think about how you extract a variable from a list.

```{r, results=hide.results()}
mylist[[2]][, "donkey"] + mylist[[3]]$BMI
```


24. Would the command mylist\$mat\$donkey + mylist\$df\$BMI work to complete task 22?

```{r, results=hide.results(), error = TRUE}
# mylist$mat$donkey + mylist$df$BMI
```

<br><br><br><br><br>

