---
title: "Practical 4.2"
author: "David Hughes"
date: "2024-10-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Practical 4.2

In this practical we are going to apply the data wrangling tools that we learned about from the second lecture of day 4 of the Introduction to R Short Course. Specifically we are going to join or merge data sets using function from the **dplyr** package andand reshape or transform other data sets using the pivot function from the **tidyr** R package. 


By the end of this practical, students should be able to:

  1.	Join data frames using dplyr.
  2.	Reshape data from wide to long and vice versa.


### Load the tidyverse compilation of packages

```{r, echo = TRUE}
library(tidyverse)
```


### The data sets:

We are going to use the **starwars** and **billboard** data set in this practical which is loaded automatically when you load the **dplyr** and  **tidyr** R packages, respectively.


### Split the starwars data set into two data sets

```{r, echo = TRUE}
set.seed(3010)
n = starwars |> pull(name) |> sample( size = 65)
n2 = starwars |> pull(name) |> sample( size = 50)
sw1 = starwars |> select(name:gender) |> filter(name %in% n)
sw2 = starwars |> select(name, homeworld:vehicles) |> filter(name %in% n2)
```

<br>


### 1. join

  - note: DO NOT call the new data set *starwars*

#### 1.1 Merge the two star wars data set into a single data set keep all the individuals present in the **sw1** data set.

  - what are the dimensions of the new data set? 
  
```{r}
mydata = left_join(sw1, sw2, by = "name")
dim(mydata)
```

#### 1.2 Merge the two star wars data set into a single data set keep all the individuals present in the **sw2** data set.

  - what are the dimensions of the new data set? 
  
```{r}
mydata2 = right_join(sw1, sw2, by = "name")
dim(mydata2)
```

#### 1.3 Merge the two star wars data sets to keep all data

  - what are the dimensions of the new data set? 
  
```{r}
mydata3 = full_join(sw1, sw2, by = "name")
dim(mydata3)
```

<br>


### 2. pivot

#### 2.1 pivot_longer() the billboard data set


  - first select() the columns from artist to wk3
  
```{r}
temp = billboard |> select(artist:wk3) |> pivot_longer(cols = starts_with("wk"), 
                                                 values_to = "count", 
                                                 names_to = "week" )

temp
```


#### 2.2 pivot_wider() the billboard data set that you just pivoted longer

```{r}
temp |> pivot_wider(names_from = week, values_from = count)
```


<br><br>

