---
title: "Simulating data for practical 2.1"
author: "PBRC: David A Hughes"
date: "Oct 16th, 2024"
output: html_document
---

In this script I am going to simulate a data set for use in the PBRC Introduction to R Short Course that will be used in practical 2.1.  

I will be simulating:

1. height
2. weight
3. sex
4. age
5. 10 proteins

### load needed libraries

```{r}
library(tidyverse)
library(faux)
```


### set a seed

```{r}
set.seed(151024)
```


### Simulate sex and age

```{r}
sex = rbinom(500, 1, 0.4)
age = rnorm(500, 35, 6)
```


### Simulate height and weight

#### height
```{r}
h = rnorm(500, 1.62, 0.08)
height = 0.13*sex + h
```


#### weight

```{r}
w = rnorm(500, 0, sqrt(1-0.33))
weight = sqrt(0.33)* as.vector( scale(height) ) + w
## make the average weight 77.5 kg (average for women) with a SD of 0.25
weight = as.vector( scale(weight) ) * 5 + 77.5
## add a sex effect
weight = 12.5*sex + weight
summary(weight)
```


```{r}
cat(paste0("Weight Summary for males\n"))
summary(weight[sex == 1])

cat(paste0("\nWeight Summary for females\n"))
summary(weight[sex == 0])
```


### estimate bmi
```{r}
bmi = weight / height^2
summary(bmi)
```


```{r}
cat(paste0("BMI Summary for males\n"))
summary(bmi[sex == 1])

cat(paste0("\nBMI Summary for females\n"))
summary(bmi[sex == 0])
```

### Simulate some molecular traits (proteins?)

```{r}
Cmat = read.table("feature_correlation_matrix.txt", header = TRUE)
Cmat = as.matrix(Cmat)
mol_traits = rnorm_multi(n = 500, vars = 10, mu = 0, sd = 1, r = Cmat)

# MT_mat = cor(mol_traits)
```

### Add some variance explained by sex, age, and BMI to the molecular traits

```{r}
x = mol_traits[,1]

## sample variance explained from a uniform distribution
sex_varexp = runif(1, min = 0, max = 0.2)
age_varexp = runif(1, min = 0, max = 0.25)
bmi_varexp = sqrt(runif(1, min = 0.05, max = 0.5))

sex_effect = sqrt(sex_varexp); if(sample(0:1, 1) == 0){ sex_effect = sex_effect * -1}
age_effect = sqrt(age_varexp); if(sample(0:1, 1) == 0){ age_varexp = age_varexp * -1}
bmi_effect = sqrt(bmi_varexp); if(sample(0:1, 1) == 0){ bmi_varexp = bmi_varexp * -1}

## add the sex, age, and bmi effects
x = sex_effect * sex + 
  age_effect * as.vector( scale(age) ) + 
  bmi_effect * as.vector( scale(bmi) ) + 
  x
x = as.vector( scale(x) )

## run a linear model
fit = lm(x ~ sex + age + bmi)

## estimate the variance explained
a = anova(fit)
estimated_etasq = ( a[,2] / sum(a[,2]) )[1:3]
varexp = data.frame(simulated = c(sex_varexp, age_varexp, bmi_varexp),
                    estimated = estimated_etasq )
# varexp
```

### write a function

```{r}
add_effects = function(x){
  ## sample variance explained from a uniform distribution
  # sex_varexp = runif(1, min = 0, max = 0.1)
  sex_varexp = rbeta(1, shape1 = 1, shape2 = 20)
  # age_varexp = runif(1, min = 0, max = 0.05)
  age_varexp = rbeta(1, shape1 = 1, shape2 = 50)
  # bmi_varexp = runif(1, min = 0, max = 0.05)
  bmi_varexp = rbeta(1, shape1 = 1, shape2 = 50)
  
  sex_effect = sqrt(sex_varexp); if(sample(0:1, 1) == 0){ sex_effect = sex_effect * -1}
  age_effect = sqrt(age_varexp); if(sample(0:1, 1) == 0){ age_effect = age_effect * -1}
  bmi_effect = sqrt(bmi_varexp); if(sample(0:1, 1) == 0){ bmi_effect = bmi_effect * -1}
  
  ## add the sex, age, and bmi effects
  x = sex_effect * sex + 
    age_effect * as.vector( scale(age) ) + 
    bmi_effect * as.vector( scale(bmi) ) + 
    as.vector( scale(x) )
  
  x = as.vector( scale(x) )
  
  ## return the variable
  return(x)

}
```

### generate new molecular traits with some random effects added

```{r}
set.seed(161024)
new_mol_traits = apply(mol_traits, 2, add_effects)
```

### test what those effects are

```{r}
e = t( sapply(1:ncol(new_mol_traits), function(i){
  x = new_mol_traits[,i]
  fit = lm(x ~ sex + age + bmi )
  coef = summary(fit)$coef
  beta = coef[2:4,1]; names(beta) = paste0("beta_", rownames(coef)[2:4])
  se = coef[2:4,2]; names(se) = paste0("se_", rownames(coef)[2:4])
  p = coef[2:4,4]; names(p) = paste0("p_", rownames(coef)[2:4])
  out = c(beta, se, p)
  return(out)
  
}) 
)

e |> knitr::kable() |> kableExtra::kable_classic()
```


## how well does the correlation structure among features hold?

```{r}
m = data.frame(old = cor(mol_traits)[,1], new = cor(new_mol_traits)[,1] )
```

## Add some missingness to the new_mol_traits

```{r}
set.seed(161024)
temp = new_mol_traits
for(i in 1:10){
  count = sample(c(0,5,10,25), 1, prob = c(0.5, 0.2, 0.15, 0.15))
  w = sample(1:nrow(temp), count)
  temp[w,i] = NA
}

apply(temp, 2, function(x){sum(is.na(x))})
```


## Define the simulated data frame

```{r}
features = as.data.frame(temp)
mydata = data.frame(sex = sex, 
                    age = age, 
                    height = height, 
                    weight = weight)
mydata = cbind(mydata, features)
```

## write the data to file

```{r}
write.table(mydata, file = "simulated_data.txt", 
            row.names = FALSE, 
            col.names = TRUE,
            sep = "\t",
            quote = FALSE)
```




